Selective Sequential Model Selection

### 2015-11-10 paper.tex
more edits

To select a model, we can feed our single-step $p$-values as inputs into sequential stopping rules such as those proposed by \citet{gsell2013sequential} and \citet{li2015accumulation}, achieving control of the familywise error rate (FWER) or false discovery rate (FDR) as desired. The FDR-controlling rules require the null $p$-values to be independent of each other and of the non-null $p$-values, a condition not satisfied by the saturated-model $p$-values of \citet{taylor2014exact}. We derive intuitive and general conditions for independence and show that our proposed constructions yield independent $p$-values. We also discuss connections between the selected-model test and parametric bootstrap sampling.
\end{abstract}

@@

To select a model, we compute a set of sequential $p$-values at each step of the model path, and then feed them into a stopping rule that is guaranteed to control the false discovery rate (FDR) \citep{benjamini1995controlling}, familywise error rate (FWER), or a similar quantity. Recently \citet{gsell2013sequential} and \citet{li2015accumulation} proposed sequential stopping rules of this kind. Both sets of rules require that once we have reached the first correct model in our path, the $p$-values in subsequent steps are uniform and independent. In this paper we develop a theoretical framework for constructing sequential $p$-values satisfying these properties, and show how to construct them in many settings.

@@

Forward stepwise regression is a greedy algorithm for building a sequence of nested linear regression models. At each iteration, it augments the current model by including the variable that will minimize the residual sum of squares of the next fitted model. Equivalently, it selects the variable with the largest $t$-statistic, controlling for the variables in the current model.

For design matrix $X\in \R^{n\times p}$ and response vector $Y\in \R^n$, let $E \sub \{1, \ldots, d\}$ denote the current {\em active set}, the set of predictor variables already selected, and let $\RSS(E)$ denote the residual sum of squares for the corresponding regression model. For $j\notin E$ let $t_{j,E}(Y)$ denote the multivariate $t$-statistic of variable $j$, adjusting for the active variables and an intercept term. Using the standard result that
\begin{equation}
t_{j,E}^2 = (n-|E|-2) \frac{\RSS(E) - \RSS(E \cup \{j\})}{\RSS(E \cup \{j\})},
\end{equation}
we see that the next variable selected is 
\begin{equation}
j^* = \argmax_{j\notin E} |t_{j,E}| = \argmin_{j\notin E} \RSS(E \cup \{j\}),
\end{equation}
with corresponding $t$-statistic $t_E^* = t_{j^*,E}$. The selective max-$t$ test rejects for large values of $t_E^*$, compared to an appropriate conditional distribution that accounts for the adaptive selection of the model path.

Table~\ref{tab:diab} illustrates the max-$t$ test and two others applied to the diabetes data from \cite{LARS}, consisting of observations of $n=442$ patients. The response of interest is a quantitative measure of disease progression one year after baseline, and there are ten measured predictors --- age, sex, body-mass index, average blood pressure, and six blood serum measurements --- plus quadratic terms, giving a total of $d=64$ features. We apply forward stepwise regression to generate the model path (beginning with an intercept term), and then use each of the three methods to generate $p$-values at each step along the path. Finally, we use the ForwardStop rule \citep{gsell2013sequential} at FDR level $\alpha=0.1$ to select a model based on the sequence of $p$-values. The red entry in each column indicates the last variable selected by ForwardStop.

@@

The nominal $p$-values in Table~\ref{tab:diab} are computed by comparing $t_E^*$ to the $t$ distribution with $n-|E|-2$ degrees of freedom, which would be the correct distribution if the sequence of models were selected before observing the data. Because the model sequence is in fact selected adaptively to maximize the value of $t_E^*$, this method is highly anti-conservative.

The max-$t$ test $p$-values are computed using the same test statistic $|t_E^*|$, but compared with a more appropriate null distribution. We can simulate the distribution of $|t_E^*|$ under the null model, i.e. the linear regression model specified by active set $E$:
\begin{equation}
  M(E):\; Y \sim N(X_E\beta, \sigma^2 I), \quad \beta\in \R^{|E|}, \sigma^2>0,
\end{equation}
where $X_E$ is the matrix with columns $(X_j)_{j\in E}$. Because $U_E=\left(X_E'Y,\, \|Y\|^2\right)$ is the complete sufficient statistic for $M(E)$, we can sample from the conditional null distribution of $t_E^*$ given $U_E$ and the current active set $E$, using the {\em selected-model test} framework of \citet{fithian2014optimal}. In step one, because $E=\emptyset$ is fixed and $t_E^*$ is independent of $U_E$, the conditional test reduces to the max-$t$ test proposed in \citet{buja2014}. In later steps, $t_E^*$ and $U_E$ are not conditionally independent given $E$, but we can approximate the null conditional distribution through Monte Carlo sampling.

@@

Section~\ref{sec:genericSetting} introduces the general problem setting, gives important definitions, and reviews relevant literature on testing ordered hypotheses using a sequence of $p$-values. Many of these methods require the $p$-values to be uniform and independent under the null, and we derive general conditions to ensure this property in Section~\ref{sec:pValsIndep}. In Section~\ref{sec:selective-reg} we contrast selected-model and saturated-model tests for linear regression setting, and illustrate why selected-model tests are often much more powerful in early steps of the path. In Section~\ref{sec:computation} we discuss strategies for sampling from the null distribution for our tests, and prove that in forward stepwise regression and $\ell_1$-regularized likelihood methods, the number of constraints in the sampling problem is never more than twice the number of variables. We provide a simulation study in Section~\ref{sec:sparseReg} and briefly discuss applications of our framework to decision trees, nonparametric changepoint detection, and principal components analysis in Section~\ref{sec:further-examples}. The paper ends with a discussion.

@@

In the general setting, we observe data $Y \in \cY$ with unknown sampling distribution $F$, and then apply some algorithm to generate an adaptive sequence of $d$ nested statistical models contained in some {\em upper model} $M_{\infty}$.
\[
M_0(Y) \sub M_1(Y) \sub \cdots \sub M_d(Y) \sub M_\infty.
\]
By {\em model}, we mean a family of candidate probability distributions for $Y$. For example in  linear regression, a model specifies  the set of
predictors allowed to have nonzero coefficients (but not the  values of their coefficients). Without loss of generality, we can take $M_\infty$ as the union of all models under consideration. We will use the notation $M_{0:d}$ to denote the sequence $M_0, \ldots, M_d$, which we call the {\em model path}. We assume throughout that for each candidate model $M$ there is a minimal sufficient statistic $U(Y; \,M)$, and write
\[
U_k(Y) = U(Y; \,M_k(Y)).
\]

@@

Define the {\em completion index} $k_0(Y) = \min\{k:\; F \in M_k(Y)\}$, the index of the first correct model. By construction, $F\in M_k \iff k \geq k_0$. A {\em stopping rule} is any estimator $\hk(Y)$ of $k_0$, with $M_k$ ``rejected'' if $k < \hk$, and ``accepted'' otherwise. Because $\hk$ is the number of models we do reject, and $k_0$ is the number we should reject, the number of type I errors is $V=(\hk-k_0)_+$, while the number of type II errors is $(k_0-\hk)_+$. Depending on the scientific context, we might want to control the FWER $\P(V>0)$, the FDR $\E[V/\hk; \hk>0]$ , or another error rate defined by the expectation of some loss function $g(\hk, k_0)$:
\begin{equation}\label{eq:errRate}
\Err_F(\hk(\cdot), g) = \E_F\left[ g\left(\hk(Y), k_0(Y)\right)\right]
\end{equation}

@@

Following \citet{fithian2014optimal}, we say that for a fixed candidate model $M$, the random variable $p_{k,M}(Y)$ is a valid {\em selective $p$-value} for $M$ at step $k$ if it is conservative under sampling from any $F\in M$, given that $M$ is selected. That is,
\begin{equation}
\P_F\left(p_{k,M}(Y) \leq \alpha \mid M_{k-1}(Y) = M\right) 
\leq \alpha, \quad \forall F\in M, \alpha \in [0,1].
\end{equation}

Once we have constructed selective $p$-values for each fixed $M$, we can use them as building blocks to construct a combined $p$-value for the random null $M_{k-1}(Y)$. Define
\[
p_k(y) = p_{k, M_{k-1}(y)}(y),
\]
which is conservative on the event $\{F \in M_{k-1}(Y)\}$:
\begin{equation}\label{eq:selectiveGuaranteePk}
\P_F\left(p_k \leq \alpha \mid F\in M_{k-1}\right) \leq \alpha, \quad \forall \alpha \in [0,1].
\end{equation}

@@

We can consider path algorithms that return a sequence of nested active sets
\[
E_0(Y) \sub E_1(Y) \sub \cdots \sub E_d(Y) \sub \cJ,
\]
inducing a model path with $M_k = M(E_k),$ for $k=0,\ldots,d$. We will be especially interested in two generic path algorithms for the sparse parametric setting: forward stepwise paths and ever-active regularization paths.

@@

\subsubsection{Ever-Active Regularization Paths and next-entry Tests}
Another important class of model selection procedures is the sequence of {\em ever-active} sets for a regularized likelihood path. 
The notion of ever-active sets is needed since these solution paths can drop  (as well as add) predictors along the way. 	For some ordered set $\Lambda$, let $\lambda\in\Lambda$ parameterize the regularization penalty $P_\lambda(\theta)$. In the case of lasso-penalized regression, we could take $P_\lambda = \lambda\|\beta\|_1$ for $\lambda\in (0,\infty]$.

@@

An example where we might not take $\ell$ to be a likelihood would be
if it were a pseudo-likelihood as in the neighborhood selection 
instead of graphical LASSO.
}
\WFcomment{Took out some of the confusing language. We are explicitly defining $\ell$ as the likelihood for $M_\infty$ at the beginning of 2.2.1. I have a remark later that $\ell$ doesn't really need to be a likelihood, but we can't accommodate completely general loss functions without losing some of the results. What we do ``need'' for the SSP is that $\ell(\theta)$ is a function of $U(Y; E)$ for $\theta \in \Theta(E)$. This is true for likelihoods but probably not true in general for pseudo-likelihoods. I'm not sure why you say this is ``really'' about $\ell_1$: it's also about elastic net, group lasso, nuclear norm, $\ell_0$, and much more. True, $\ell_1$ is the prototypical example, but I don't see how we get anything from explicitly restricting ourselves to $\ell_1$.}

Assuming for simplicity that there is a unique solution to each penalized problem of the form
\begin{equation}\label{eq:pen-lik}
  \hat\theta^{\lambda}(Y) =
  \argmin_{\theta\in\Theta} -\ell(\theta; Y) + P_\lambda(\theta),
\end{equation}
we define the ever-active set for $\lambda\in\Lambda$ as
\begin{equation}
  \tE_\lambda(Y) = \left\{j:\; \hat\theta_j^\nu(Y) \neq 0 
    \text{ for any } \nu\in \Lambda, \nu \geq \lambda \right\}
\end{equation}
It may be impossible or inconvenient to compute $\hat\theta^\lambda(Y)$ for every $\lambda\in(0,\infty]$. If so, we can instead take $\Lambda$ to be a grid of finitely many values.

@@

\paragraph{Remark} In its usual formulation, the lasso coefficients for linear regression minimize a penalized RSS criterion:
\[
\hat\beta_{\text{lasso}}^\lambda = \frac{1}{2}\|Y-X\beta\|_2^2 + \lambda\|\beta\|_1.
\]

This is slightly different from the penalized log-likelihood criterion of \eqref{eq:pen-lik}, which would minimize
\begin{equation}\label{eq:lasso-beta-sigma}
  -\ell(\beta, \sigma;\,Y) + P_\lambda(\beta)=
  \frac{n}{2}\log(2\pi\sigma^2) + \frac{1}{2\sigma^2}\|Y-X\beta\|_2^2 
  + \lambda\|\beta\|_1.
\end{equation}

@@

As a result the lasso coefficient vector $\hat\beta_{\text{lasso}}^\lambda$ minimizes
\[
\frac{n}{2\pi e}\exp\left\{-2\ell^*(\beta;\,Y)/n\right\} + \lambda\|\beta\|_1,
\]
while the regularized log-likelihood estimator $\hat\beta^\lambda$ minimizes $-\ell^*(\beta;\,Y)+\lambda\|\beta\|_1$. If $\Lambda=(0,\infty)$ then both methods result in exactly the same ever-active path. Otherwise, at the cost of some notational bloat, we could generalize \eqref{eq:pen-lik} by replacing $-\ell(\cdot)$ with $\rho(\ell(\cdot))$, where $\rho$ is any strictly decreasing function, and the theory in this article would go through unchanged.

\begin{comment}
\begin{align*}
  \hat\sigma^2(\beta)
  &\triangleq \argmax_{\sigma^2>0} \ell(\beta,\sigma^2;\,Y) 
  = \frac{1}{n}\|Y-X\beta\|_2^2, \quad \text{ and }
  \\
  \ell^*(\beta; \,Y)
  &\triangleq \sup_{\sigma^2>0} \ell(\beta, \sigma^2;\,Y) 
  = -\frac{n}{2} - \frac{n}{2}\log\left(\frac{2\pi\|Y-X\beta\|_2^2}{n}\right).
\end{align*}

As a result the lasso coefficients $\hat\beta_{\text{lasso}}^\lambda$ minimize
\[
\exp\{-2\ell(\beta,\sigma;\,Y)/n \,+\, \text{const.}\} + \lambda\|\beta\|_1,
\]
instead of the criterion in \eqref{eq:pen-lik}. If $\Lambda=(0,\infty)$ then both methods result in exactly the same ever-active path. Otherwise, we could ``fix'' this issue (at the cost of some notational bloat) by replacing $-\ell(\cdot)$ with $\rho(\ell(\cdot))$, where $\rho$ is any strictly decreasing function, and all of the theory in this article would go through unchanged.
\end{comment}

@@

Note that in our formulation of the problem, the type I error $V=(\hk-k_0)_+$ is defined in a ``model-centric'' fashion: at step $k$, we are testing  whether a particular linear model $M(E_{k-1})$ adequately describes the data $Y$. Even if the next selected variable $X_{j_k}$ is a noise variable in the full model, it is not necessarily a mistake to reject $M(E_{k-1})$ provided there are other signal variables that have not yet been included.

In some circumstances, we might want to define a type I error at step $k$ differently, by choosing a different null hypothesis to test. Let $\mu = \E Y$ and let $\theta^E$ denote the {\em least-squares coefficients} of active set $E$ --- the coefficients of the best linear predictor for the design matrix $X_E$:
\[
\theta^E = X_E^+ \mu = \argmin_{\theta\in \R^{|E|}} \|\mu - X_E\theta\|_2^2,
\]
where $X_E^+$ is the Moore-Penrose pseudoinverse of the matrix $X_E$. 
These are the true ``partial regression coefficients''  for design matrix $X_E$.
 Write $\proj_E\mu = X_E \theta^E$ for the projection of $\mu$ into the column space of $X_E$.

@@

In the first procedure, we construct a (type I) analysis-of-variance (ANOVA) table performing an $F$- or $t$-test of each model against the next. We then stop and select model $M_k=M(\{1,\ldots,k\})$ the first time the $F$-test $p$-value is larger than $\alpha$. This procedure would control the complete-null and incremental-null FWER, but {\em not} the full-model-null FWER. The methods we consider in this article are essentially generalizations of this procedure for the setting where the model sequence is selected adaptively, and where we may wish to control FDR instead of FWER.

@@

\section{Stopping Rules, Ordered Testing, and Independence}

@@

Since $k_0$ is the index of the first null hypothesis, we have
\begin{equation}\label{eq:basic-stop-control}
 \text{FWER} \leq \P(p_{k_0} \leq \alpha) \leq alpha.
\end{equation}

@@

\citet{gsell2013sequential} propose the {\em ForwardStop} rule to controls the FDR:
\[
  \hk_{F}(Y) = \max\left\{k:\;
    -\frac{1}{k}\sum_{i=1}^k \log(1-p_i) \leq \alpha\right\}
\]
ForwardStop uses the $p$-values {\it before} step $k$ in making its decision whether to reject at step $k$. If $p_k$ is uniform, then $-\log(1-p_k)$ is an $\text{Exp}(1)$ random variable with expectation 1; thus, the sum can be seen as an estimate of the false discovery proportion (FDP):
\[
\widehat{\text{FDP}}_k = -\frac{1}{k}\sum_{i=1}^k \log(1-p_i).
\]
ForwardStop chooses the largest model for which $\widehat{\text{FDP}}_k \leq \alpha$. \citet{gsell2013sequential} show that ForwardStop controls the FDR if the null $p$-values are independent of each other and of the non-nulls.

@@

In our problem setting, the sequence of selected models is random; thus, the truth or falsehood of each $H_k$ is not fixed, as assumed in the previous section. It is not immediately clear, then, how much we should condition on when we construct the $p$-value at step $k$. In particular, is it sufficient to condition only on the current null model $M_{k-1}$, or must we condition on the entire subpath of models $M_0,\ldots,M_{k-1}$?

Note that conditioning on the current null model {\em is} enough to guarantee that $p_k$ is uniform conditional on the event $\{F\in M_{k-1}(Y)\}$ that the $k$th null is true. Unfortunately, however, it is {\em not} enough to guarantee that the above stopping rules actually control FDR or FWER.

Let $z_{\alpha/2}$ denote the upper-$\alpha/2$ quantile of the $\cN(0,1)$ distribution and consider the following counterexample:

\begin{restatable}{proposition}{counterex}\label{prop:counterexample}
  Consider linear regression with $n=p=3$, known $\sigma^2=1$ and identity design $X=I_3$.

  Suppose that we construct the model path as follows: if $|Y_3|>z_{\alpha/2}$, we add $X_1$ to the active set first, then $X_2$, then $X_3$. Otherwise, we add $X_2$, then $X_1$, then $X_3$. At each step we construct selective max-$z$ $p$-values and finally choose $\hat k$ using BasicStop.

  If $\mu=(0,C,0)$, then the FWER for the entire procedure tends to $2\alpha-\alpha^2$ as $C\to\infty$.
\end{restatable}

@@

If, however, we could guarantee that
\[
\P(p_k \leq \alpha \mid H_k \text{ true}, k_0=k) \leq \alpha
\]
then BasicStop again would control FWER, by the argument in \ref{eq:basic-stop-control}.

>>>>

To select a model, we can feed our single-step $p$-values as inputs into sequential stopping rules such as those proposed by \citet{gsell2013sequential} and \citet{li2015accumulation}, achieving control of the familywise error rate or false discovery rate (FDR) as desired. The FDR-controlling rules require the null $p$-values to be independent of each other and of the non-null $p$-values, a condition not satisfied by the saturated-model $p$-values of \citet{taylor2014exact}. We derive intuitive and general sufficient conditions for independence and show that our proposed constructions yield independent $p$-values.
\end{abstract}

@@

To select a model, we compute a set of sequential $p$-values at each step of the model path, and then feed them into a stopping rule that is guaranteed to control the false discovery rate (FDR) \citep{benjamini1995controlling}, familywise error rate (FWER), or a similar quantity. Recently \citet{gsell2013sequential} and \citet{li2015accumulation} proposed sequential stopping rules of this kind. Both sets of rules require that once we have reached the first correct model in our path, the $p$-values in subsequent steps are uniform and independent. In this paper we develop a theoretical framework for constructing sequential $p$-values satisfying these properties, and give explicit constructions.

@@

Forward stepwise regression is a greedy algorithm for building a sequence of nested linear regression models. At each iteration, it augments the current model by including the variable that will minimize the residual sum of squares (RSS) of the next fitted model. Equivalently, it selects the variable with the largest $t$-statistic, adjusting for the variables in the current model.

For design matrix $X\in \R^{n\times p}$ and response vector $Y\in \R^n$, let $E \sub \{1, \ldots, d\}$ denote the current {\em active set}, the set of predictor variables already selected, and let $\RSS(E)$ denote the residual sum of squares for the corresponding regression model. For $j\notin E$ let $t_{j,E}(Y)$ denote the multivariate $t$-statistic of variable $j$, adjusting for the active variables as well as an intercept term. Using the standard result that
\begin{equation}
t_{j,E}^2 = (n-|E|-2) \frac{\RSS(E) - \RSS(E \cup \{j\})}{\RSS(E \cup \{j\})},
\end{equation}
we see that the next variable selected is 
\begin{equation}
j^* = \argmax_{j\notin E} |t_{j,E}| = \argmin_{j\notin E} \RSS(E \cup \{j\}),
\end{equation}
with corresponding $t$-statistic $t_E^* = t_{j^*,E}$. The selective max-$t$ test rejects for large values of $|t_E^*|$, compared to an appropriate conditional null distribution that accounts for the adaptive selection of the model path.

Table~\ref{tab:diab} illustrates the max-$t$ test and two others applied to the diabetes data from \cite{LARS}, consisting of observations of $n=442$ patients. The response of interest is a quantitative measure of disease progression one year after baseline, and there are ten measured predictors --- age, sex, body-mass index, average blood pressure, and six blood serum measurements --- plus quadratic terms, giving a total of $d=64$ features. We apply forward stepwise regression to generate the model path (beginning with an intercept term), and then use each of the three methods to generate $p$-values at each step along the path. Finally, we use the ForwardStop rule \citep{gsell2013sequential} at FDR level $\alpha=0.1$ to select a model based on the sequence of $p$-values. The bolded entry in each column indicates the last variable selected by ForwardStop.

@@

The nominal $p$-values in Table~\ref{tab:diab} are computed by comparing $t_E^*$ to the $t$ distribution with $n-|E|-2$ degrees of freedom, which would be the correct distribution if the sequence of models were selected before observing the data. Because the model sequence is in fact selected adaptively to maximize the absolute value of $t_E^*$, this method is highly anti-conservative.

The max-$t$ test $p$-values are computed using the same test statistic $|t_E^*|$, but compared with a more appropriate null distribution. We can simulate the distribution of $|t_E^*|$ under the null model, i.e. the linear regression model specified by active set $E$:
\begin{equation}
  M(E):\; Y \sim \cN(X_E\beta_E, \sigma^2 I), \quad \beta_E\in \R^{|E|}, \sigma^2>0,
\end{equation}
where $X_E$ is the matrix with columns $(X_j)_{j\in E}$. Because $U_E=\left(X_E'Y,\, \|Y\|^2\right)$ is the complete sufficient statistic for $M(E)$, we can sample from the conditional null distribution of $t_E^*$ given $U_E$ and the current active set $E$, using the {\em selected-model test} framework of \citet{fithian2014optimal}. In step one, because $E=\emptyset$ is fixed and $t_E^*$ is independent of $U_\emptyset=\|Y\|^2$, the conditional test reduces to the max-$t$ test proposed in \citet{buja2014}. In later steps, $t_E^*$ and $U_E$ are not conditionally independent given $E$, but we can numerically approximate the null conditional distribution through Monte Carlo sampling.

@@

Sections~\ref{sec:genericSetting}--\ref{sec:ordered} introduce the general problem setting, give important definitions, and review relevant literature on testing ordered hypotheses using a sequence of $p$-values. Many of these methods require the $p$-values to be uniform and independent under the null, and we derive general conditions to ensure this property in Section~\ref{sec:pValsIndep}. In Section~\ref{sec:selective-reg} we contrast selected-model and saturated-model tests in the linear regression setting, and explain why selected-model tests are often much more powerful in early steps of the path. In Section~\ref{sec:computation} we discuss strategies for sampling from the null distribution for our tests, and prove that in forward stepwise regression and $\ell_1$-regularized likelihood methods, the number of constraints in the sampling problem is never more than twice the number of variables. We provide a simulation study in Section~\ref{sec:sparseReg}, and Section~\ref{sec:further-examples} discusses applications of our framework beyond linear regression to decision trees nonparametric changepoint detection. The paper ends with a discussion.

@@

In the general setting, we observe data $Y \in \cY$ with unknown sampling distribution $F$, and then apply some algorithm to generate an adaptive sequence of $d$ nested statistical models contained in some {\em upper model} $M_{\infty}$:
\[
M_0(Y) \sub M_1(Y) \sub \cdots \sub M_d(Y) \sub M_\infty.
\]
By {\em model}, we mean a family of candidate probability distributions for $Y$. For example in  linear regression, a model specifies  the set of
predictors allowed to have nonzero coefficients (but not the  values of their coefficients). Note the ``assumption'' of an upper model involves no loss of generality --- we can always take $M_\infty$ as the union of all models under consideration. We will use the notation $M_{0:d}$ to denote the sequence $M_0, \ldots, M_d$, which we call the {\em model path}. We assume throughout that for each candidate model $M$ there is a minimal sufficient statistic $U(Y; \,M)$, and write
\[
U_k(Y) = U(Y; \,M_k(Y)).
\]

@@

Define the {\em completion index} $k_0(Y) = \min\{k:\; F \in M_k(Y)\}$, the index of the first correct model. By construction, $F\in M_k \iff k \geq k_0$. A {\em stopping rule} is any estimator $\hk$ of $k_0$, with $M_k$ ``rejected'' if $k < \hk$, and ``accepted'' otherwise. Because $\hk$ is the number of models we do reject, and $k_0$ is the number we should reject, the number of type I errors is $V=(\hk-k_0)_+$, while the number of type II errors is $(k_0-\hk)_+$. Depending on the scientific context, we might want to control the FWER $\P(V>0)$, the FDR $\E[V/\hk; \hk>0]$ , or another error rate such as a modified FDR, defined by the expectation of some loss function $g(\hk, k_0)$:
\begin{equation}\label{eq:errRate}
\Err_F(\hk(\cdot), g) = \E_F\left[ g\left(\hk(Y), k_0(Y)\right)\right]
\end{equation}

@@

Following \citet{fithian2014optimal}, we say that for a fixed candidate null model $M$, the random variable $p_{k,M}(Y)$ is a valid {\em selective $p$-value} for $M$ at step $k$ if it is stochastically larger than uniform (henceforth super-uniform) under sampling from any $F\in M$, given that $M$ is selected. That is,
\begin{equation}
\P_F\left(p_{k,M}(Y) \leq \alpha \mid M_{k-1}(Y) = M\right) 
\leq \alpha, \quad \forall F\in M, \alpha \in [0,1].
\end{equation}

Once we have constructed selective $p$-values for each fixed $M$, we can use them as building blocks to construct a combined $p$-value for the random null $M_{k-1}(Y)$. Define
\[
p_k(y) = p_{k, M_{k-1}(y)}(y),
\]
which is super-uniform on the event $\{F \in M_{k-1}(Y)\}$:
\begin{equation}\label{eq:selectiveGuaranteePk}
\P_F\left(p_k \leq \alpha \mid F\in M_{k-1}\right) \leq \alpha, \quad \forall \alpha \in [0,1].
\end{equation}

@@

We can consider path algorithms that return a sequence of nested active sets
\[
E_0(Y) \sub E_1(Y) \sub \cdots \sub E_d(Y) \sub \cJ,
\]
inducing a model path with $M_k = M(E_k),$ for $k=0,\ldots,d$. We will be especially interested in two generic path algorithms for the sparse parametric setting: forward stepwise paths and ever-active regularization paths. For a nonparametric example see Section~\ref{sec:further-examples}

@@

\subsubsection{Ever-Active Regularization Paths and Next-Entry Tests}
Another important class of model selection procedures is the sequence of {\em ever-active} sets for a regularized likelihood path. 
The notion of ever-active sets is needed since these solution paths can drop  (as well as add) predictors along the way. 	For some ordered set $\Lambda$, let $\lambda\in\Lambda$ parameterize the regularization penalty $P_\lambda(\theta)$. In the case of lasso-penalized regression, we could take $P_\lambda = \lambda\|\beta\|_1$ for $\lambda\in (0,\infty)$.

@@

An example where we might not take $\ell$ to be a likelihood would be
if it were a pseudo-likelihood as in the neighborhood selection 
instead of graphical LASSO.
}
\WFcomment{Took out some of the confusing language. We are explicitly defining $\ell$ as the likelihood for $M_\infty$ at the beginning of 2.2.1. I have a remark later that $\ell$ doesn't really need to be a likelihood, but we can't accommodate completely general loss functions without losing some of the results. What we do ``need'' for the SSP is that $\ell(\theta)$ is a function of $U(Y; E)$ for $\theta \in \Theta(E)$. This is true for likelihoods but probably not true in general for pseudo-likelihoods. I'm not sure why you say this is ``really'' about $\ell_1$: it's also about elastic net, group lasso, nuclear norm, $\ell_0$ (such as in PCA), and much more. True, $\ell_1$ is the prototypical example, but I don't see how we get anything from explicitly restricting ourselves to $\ell_1$.}

Assume for simplicity that there is a unique solution to each penalized problem of the form
\begin{equation}\label{eq:pen-lik}
  \hat\theta^{\lambda}(Y) =
  \argmin_{\theta\in\Theta} -\ell(\theta; Y) + P_\lambda(\theta).
\end{equation}
It may be impossible or inconvenient to compute $\hat\theta^\lambda(Y)$ for every $\lambda\in(0,\infty)$. If so, we can instead take $\Lambda$ to be a grid of finitely many values. We define the {\em ever-active set} for $\lambda\in\Lambda$ as
\begin{equation}
  \tE_\lambda(Y) = \left\{j:\; \hat\theta_j^\nu(Y) \neq 0 
    \text{ for any } \nu\in \Lambda, \nu \geq \lambda \right\}
\end{equation}

@@

\paragraph{Remark} In its usual formulation, the lasso coefficients for linear regression minimize a penalized RSS criterion:
\[
\hat\beta_{\text{lasso}}^\lambda = \frac{1}{2}\|Y-X\beta\|_2^2 + \lambda\|\beta\|_1,
\]
which is slightly different from the penalized log-likelihood criterion of \eqref{eq:pen-lik}. One can show that the lasso coefficients $\hat\beta_{\text{lasso}}^\lambda$, along with $\hat\sigma^2=n^{-1}\|Y-X\hat\beta^\lambda_{\text{lasso}}\|^2$, minimize
\[
\frac{n}{2\pi e}\exp\left\{-2\ell(\beta, \sigma^2;\,Y)/n\right\} + \lambda\|\beta\|_1.
\]
If $\Lambda=(0,\infty)$ then both methods result in exactly the same ever-active path. Otherwise, at the cost of some notational bloat, we could generalize \eqref{eq:pen-lik} by replacing $-\ell(\cdot)$ with $\rho(\ell(\cdot))$ for any strictly decreasing $\rho$, and the theory in this article would go through unchanged.

\begin{comment}
This is slightly different from the penalized log-likelihood criterion of \eqref{eq:pen-lik}, which would minimize
\begin{equation}\label{eq:lasso-beta-sigma}
  -\ell(\beta, \sigma;\,Y) + P_\lambda(\beta)=
  \frac{n}{2}\log(2\pi\sigma^2) + \frac{1}{2\sigma^2}\|Y-X\beta\|_2^2 
  + \lambda\|\beta\|_1.
\end{equation}

@@

As a result the lasso coefficient vector $\hat\beta_{\text{lasso}}^\lambda$ minimizes
\[
\frac{n}{2\pi e}\exp\left\{-2\ell^*(\beta;\,Y)/n\right\} + \lambda\|\beta\|_1,
\]
while the regularized log-likelihood estimator $\hat\beta^\lambda$ minimizes $-\ell^*(\beta;\,Y)+\lambda\|\beta\|_1$. If $\Lambda=(0,\infty)$ then both methods result in exactly the same ever-active path. Otherwise, at the cost of some notational bloat, we could generalize \eqref{eq:pen-lik} by replacing $-\ell(\cdot)$ with $\rho(\ell(\cdot))$, where $\rho$ is any strictly decreasing function, and the theory in this article would go through unchanged.
\end{comment}

\begin{comment}
\begin{align*}
  \hat\sigma^2(\beta)
  &\triangleq \argmax_{\sigma^2>0} \ell(\beta,\sigma^2;\,Y) 
  = \frac{1}{n}\|Y-X\beta\|_2^2, \quad \text{ and }
  \\
  \ell^*(\beta; \,Y)
  &\triangleq \sup_{\sigma^2>0} \ell(\beta, \sigma^2;\,Y) 
  = -\frac{n}{2} - \frac{n}{2}\log\left(\frac{2\pi\|Y-X\beta\|_2^2}{n}\right).
\end{align*}
\end{comment}

@@

Note that in our formulation of the problem, the type I error $V=(\hk-k_0)_+$ is defined in a ``model-centric'' fashion: at step $k$ in linear regression, we are testing  whether a particular linear model $M(E_{k-1})$ adequately describes the data $Y$. Even if the next selected variable $X_{j_k}$ is a noise variable in the full model, it is not necessarily a mistake to reject $M(E_{k-1})$ provided there are other signal variables that have not yet been included.

In some circumstances, we might want to define a type I error at step $k$ differently, by choosing a different null hypothesis to test. Let $\mu = \E Y$ and let $\theta^E$ denote the {\em least-squares coefficients} of active set $E$ --- the coefficients of the best linear predictor for the design matrix $X_E$:
\[
\theta^E = X_E^+ \mu = \argmin_{\theta\in \R^{|E|}} \|\mu - X_E\theta\|_2^2,
\]
where $X_E^+$ is the Moore-Penrose pseudoinverse of the matrix $X_E$.

@@

In the first procedure, we construct a (type I) analysis-of-variance (ANOVA) table performing an $F$- or $t$-test of each model against the next. We then stop and select model $M_k=M(\{1,\ldots,k\})$ the first time the $F$-test $p$-value is larger than $\alpha$. This procedure would control the complete-null and incremental-null FWER, but {\em not} the full-model-null FWER. The methods we consider in this article are essentially generalizations of this procedure to the setting where the model sequence is selected adaptively, and where we may wish to control FDR instead of FWER.

@@

\section{Stopping Rules and Ordered Testing}\label{sec:ordered}

@@

Since $k_0$ is the index of the first null hypothesis, we have
\begin{equation}\label{eq:basic-stop-control}
 \text{FWER} \leq \P(p_{k_0} \leq \alpha) \leq \alpha.
\end{equation}

@@

\citet{gsell2013sequential} propose the {\em ForwardStop} rule to control the FDR:
\[
  \hk_{F}(Y) = \max\left\{k:\;
    -\frac{1}{k}\sum_{i=1}^k \log(1-p_i) \leq \alpha\right\}
\]
If $p_k$ is uniform, then $-\log(1-p_k)$ is an $\text{Exp}(1)$ random variable with expectation 1; thus, the sum can be seen as an estimate of the false discovery proportion (FDP):
\[
\widehat{\text{FDP}}_k = -\frac{1}{k}\sum_{i=1}^k \log(1-p_i),
\]
and we choose the {\em largest} model with $\widehat{\text{FDP}}_k \leq \alpha$. \citet{gsell2013sequential} show that ForwardStop controls the FDR if the null $p$-values are independent of each other and of the non-nulls.

@@

In our problem setting, the sequence of selected models is random; thus, the truth or falsehood of each $H_k$ is not fixed, as assumed in the previous section. It is clear that we will recover all of the same guarantees if we construct our $p$-values conditional on the entire path $M_{0:d}(Y)$; however, this option is rather unappealing for both computational and statistical reasons. In constructing the $p$-value for step $k$, is it sufficient to condition only on the subpath of models $M_{0:k-1}(Y)$, or better yet to condition only on the current null model $M_{k-1}$?

Note that conditioning on the current null model {\em is} enough to guarantee that $p_k$ is uniform conditional on the event $\{F\in M_{k-1}(Y)\}$ that the $k$th null is true. Unfortunately, however, it is {\em not} enough to guarantee that the above stopping rules actually control FDR or FWER. Let $z_{\alpha/2}$ denote the upper-$\alpha/2$ quantile of the $\cN(0,1)$ distribution, and consider the following counterexample:

\begin{restatable}{proposition}{counterex}\label{prop:counterexample}
  Consider linear regression with $n=p=3$, known $\sigma^2=1$ and identity design $X=I_3$, and suppose that we construct the model path as follows: if $|Y_3|>z_{\alpha/2}$, we add $X_1$ to the active set first, then $X_2$, then $X_3$. Otherwise, we add $X_2$, then $X_1$, then $X_3$. At each step we construct selective max-$z$ $p$-values and finally choose $\hat k$ using BasicStop.

  If $\mu=(0,C,0)$, then the FWER for this procedure tends to $2\alpha-\alpha^2$ as $C\to\infty$.
\end{restatable}

@@

If, however, we could guarantee that
\[
\P(p_k \leq \alpha \mid H_k \text{ true},\, k_0=k) \leq \alpha
\]
then BasicStop again would control FWER, by the argument in \ref{eq:basic-stop-control}.
